Projit: An Open Source tool for Decoupled Data Science

### 2023-06-26 docs/paper/paper.tex
Paper update

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}. 
To do so they are often constructed for specific combinations of technology. This is a necessary limitation 
to enable efficiency, but has the effect of limiting general applicability.
Many other eScience frameworks focus on the lineage and management of data, referred to as 
the so-called provenance problem \cite{Sahoo:2008,Conquest:2021}
The goal of the provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple source systems,
or it can be auditing of logs generated during data processing\cite{Ferdous2020}. Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and processes so
that auditing can be easily performed over many projects.

In addition to systems for storage of data, eScience application may include facilities for orchestration
of data processes and services\cite{Subramanian2013}, analysis of results, generation of insights
and documentation.

Frameworks for eScience will typically need to take a position on the extent to which they are domain
specific, versus general purpose. A domain specific approach that integrates multiple data sources in
a domain aware fashion can faciliate automated or assisted scientific discover\cite{Howe2008}. On the
other hand a general purpose framework facilitates multi-disciplinary collaboration and permits meta-analysis
that transcends the boundaries of disciplines.

Other frameworks and approaches in eScience focus on understanding how to do large scale collaborative science, or
facilitate meta-level learning of various kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve scientific processes
as well as data mine the history of science for phenomena that were difficult to detect.

@@

In Figure \ref{fig:projit} we see that the core steps of data preparation, experimentation
and analysis of results all happen independently. Each of them accesses the projit store for
the information they need, storing information

>>>>

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}. 
To do so they are often constructed for specific combinations of technology. This is a necessary limitation 
to enable efficiency, but has the effect of limiting general applicability.
Many other eScience frameworks focus on the lineage and management of data, referred to as 
the so-called provenance problem \cite{Sahoo:2008,Conquest:2021}
The goal of the provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple source systems,
auditing of logs generated during data processing\cite{Ferdous2020}. At the extreme we can seek to
quantify every transformation that happens to data in the course of processing\cite{Sahoo2009}. 
Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and processes so
that auditing can be easily performed over many projects.

In addition to systems for storage of data, eScience application may include facilities for orchestration
of data processes and external services\cite{Subramanian2013}, requests for experiments with specific 
parameters\cite{Hunter:2005}, or integrated analysis of results, generation of insights and documentation. 
Other frameworks and approaches in eScience focus on understanding how to do large scale collaborative science, or
facilitate meta-level learning of various kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve scientific processes
as well as data mine the history of experimental results for phenomena that were difficult to detect.

Frameworks for eScience will typically need to take a position on the extent to which they are domain
specific, versus general purpose. A domain specific approach that integrates multiple data sources in
a domain aware fashion can faciliate automated or assisted scientific discover\cite{Howe2008}. On the
other hand a general purpose framework facilitates multi-disciplinary collaboration and permits meta-analysis
that transcends the boundaries of disciplines. The other key dimension for a decision is the extent to
which an eScience application depends on specific technologies, many machine learning science platforms
can provide efficiency gains, but only when using specfic libraries and frameworks\cite{Alberti:2018,MolnerDomenech:2020}.
Similarly, other empirical science platforms are built on specific database, webserver or application
frameworks, which make them less extensible and harder to integrate.

In this work we argue for development of data science frameworks that are minimal in expectations, both in
terms of appplication domains and underlying technologies. We present a design framework for building
decoupled data science tools that can improve efficiency and replication through standardisation, without
unreasonable impositions on design decisions. We describe the design of a open source project integration
tool (\textit{projit}) that can be used either as a CLI or python API. Internally \textit{projit} depends 
only on a metadata store that uses the general purpose JSON format. 
As such it is trivial for developers to build interfaces in other
languages, or devise web service APIs for decentralised versions. We explore a case study of comparing results
across multiple projects for which we have used the \textit{projit} application to manage our metadata.

@@

In Figure \ref{fig:projit} we see that the core steps of data preparation, experimentation
and analysis of results all happen independently. Each of them accesses the projit metadata 
store for the information they need, and subsequently store information and results once complete.
This process means that in principle the location of an underlying dataset could change without
modifying other elements of the project. Similarly, we might change the parameters or an experiment
or the set of metrics we calculate. Each experiment and analysis task operates independently of the
others, and all that suffers when changes are made is the potential for comparison across equal
dimensions of variation.

In addition to the dominant requirements of experimentation (parameter and results) we store the results
of each experimental execution as well as the experiment duration, measured from the time of initiation
to completion. These records are particularly important in data science and machine learning where we
may want to trade off performance with computational requirements. But these values could be used to store
information about real world experimtal execution, or the time required to marshal and sequence a series
of independent web services.
